### 2017-01-03 Chapters/theory.tex
Change titlepage (new format) and write Suppinfo S5 about strains
Correct some figures and add the suppinfo of chapter 3 except the part about the results obtained in microplate reader
First draft of summary of chapter 2

\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies }
\chaptermark{Dynamical allocation of cell resources}
\label{chap:theory}

@@

%\section*{Summary of chapter \thechapter}
%Microbial physiology exhibits growth laws that relate the macromolecular composition of the cell to the growth rate.
%Recent work has shown that these empirical regularities can be derived from coarse-grained models of resource allocation.
%While these studies focus on steady-state growth, such conditions are rarely found in natural habitats, where microorganisms are continually challenged by environmental fluctuations.
%The aim of this paper is to extend the study of microbial growth strategies to dynamical environments, using a self-replicator model.
%We formulate dynamical growth maximization as an optimal control problem that can be solved using Pontryagin’s Maximum Principle.
%We compare this theoretical gold standard with different possible implementations of growth control in bacterial cells.
%We find that simple control strategies enabling growth-rate maximization at steady state are suboptimal for transitions from one growth regime to another, for example when shifting bacterial cells to a medium supporting a higher growth rate.
%A near-optimal control strategy in dynamical conditions is shown to require information on several, rather than a single physiological variable.
%Interestingly, this strategy has structural analogies with the regulation of ribosomal protein synthesis by ppGpp in the enterobacterium \textit{Escherichia coli}.
%It involves sensing a mismatch between precursor and ribosome concentrations, as well as the adjustment of ribosome synthesis in a switch-like manner.
%Our results show how the capability of regulatory systems to integrate information about several physiological variables is critical for optimizing growth in a changing environment.

\selectlanguage{french}
\section*{Résumé en français du chapître \thechapter : [titre français]}
dans cette section blablabla

\selectlanguage{english}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=0.9]{./Fig/Fig1.eps}
\caption{
\textbf{Self-replicator model of bacterial growth.}
External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

\begin{figure}[tb]
\includegraphics[width=\textwidth]{./Fig/Fig2.eps}
\caption{\textbf{Analysis of self-replicator model of bacterial growth.}
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{./Fig/Fig3.eps}
\caption{
\textbf{Self-replicator model accounts for bacterial growth laws.}
\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
\label{fig:model_validation}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig4.eps}
\caption{
\textbf{Optimal control of the self-replicator during a nutrient upshift.}
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig5.eps}
\caption{
\textbf{Alternative strategies for controlling the self-replicator of bacterial growth.}
The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
}
\label{fig:strategies_overview}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig6.eps}
\caption{
\textbf{Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig7.eps}
\caption{
\textbf{Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=1]{./Fig/Fig8.eps}
\caption{
\textbf{ppGpp regulation implements an on-off control strategy of resource allocation.}
\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
}
\label{fig:ppGppsurface}
\end{figure}

@@

%%%%%% FOR MANUAL REF%%%%%
\makeatletter
\newcommand{\manuallabel}[2]{\def\@currentlabel{#2}\label{#1}}
\makeatother
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding.eps}
\caption
{
\textbf{Local stability of the on-off strategy.}
The on-off strategy sets $\alpha$ to a value of 0 (1) when $\hat{r} > g(\hat{p})$ ($\hat{r} < g(\hat{p})$).
The solid, black curve is the $\hat{p}$-nullcline.
The dashed, black curve is the curve $\hat{r}=g(\hat{p})$.
The arrows represent the vector fields for $\alpha=0$ (in blue) and $\alpha=1$ (in red).
The intersection of the $\hat{p}$-nullcline and the curve $\hat{r}=g(\hat{p})$ corresponds to a unique non-trivial stable steady state, which is equal to $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ by Eq.~\ref{eq:supp_adim_h}.
}
\label{fig:sliding}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[height=4cm]{Fig/adj2ab.eps} 
\includegraphics[height=4cm]{Fig/adj2.eps} 
\caption{\textbf{Transitions between regions in the phase-plane for the adjoint system.} A switch occurs when a trajectory crosses the $\lambda_p$-axis. Left: abnormal case. An extremal trajectory cannot have more than two switches. Right: normal case. $(\lambda_p^s,0)$ corresponds to the singular arc.  After the first switch, an extremal trajectory cannot have two consecutive switches if it stays in the region $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ or $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$.}
\label{fig-adj}                               
\end{figure}

@@

\begin{figure}[tb]~Text
\centering
\includegraphics[width=0.7\textwidth]{./Fig/FigS4-1.eps}
\caption[ppGpp concentration is a function of total ribosome and amino acid concentrations.]
{
{\bf ppGpp concentration is a function of total ribosome and amino acid concentrations.}\newline
We assume the dynamics of ppGpp to be fast on the time-scale of changes in the ribosome and amino acid concentrations.
The concentration of ppGpp can thus be expressed as a function of the latter two variables, using the model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}.
Parameters are taken from Table~\ref{tab:notations}.
}
\label{fig:ppGpp}
\end{figure}

\begin{table}[tb]
\scriptsize
\centering
\begin{tabular}{cccl}
\hline
\hline
\textbf{Symbol} & \textbf{Value} & \textbf{Unit} & \textbf{Description} \\
\hline
\hline
$a_i$ & -- & $\mu$M & Concentration of aa $i$ (not incorporated in protein)\\

>>>>

\chapter{Dynamical Allocation of Cellular Resources as an Optimal Control Problem: Novel Insights into Microbial Growth Strategies}
\chaptermark{Dynamical allocation of cell resources}
\label{chap:theory}

@@

\begin{center}
\noindent\rule{4cm}{0.1pt}
\end{center}

\selectlanguage{french}
\begin{chapter_summary}{Modélisation de l'allocation dynamique des ressources cellulaires comme un problème de contrôle optimal}

Ce chapitre est dédié à une approche théorique du problème d'allocation des ressources cellulaires chez les microorganismes.
En Section \ref{sec:model}, mous modélisons la cellule comme un auto-réplicateur (Fig.~\ref{fig:self_replicator}) devant allouer une ressource commune (les nutriments du milieu) à deux secteurs distincts de macromolécules: la machinerie métabolique qui extrait les nutriments de l'environnement et les convertit en précurseurs utilisables, et la machinerie d'expression génique qui utilise ces précurseurs pour produire de nouvelles macromolécules.
Ce modèle très simple comprenant 2 réactions (Eq.~\ref{eq:reactions}) se traduit mathématiquement par deux équations différentielles ordinaires (Eqs~\ref{eq:pdef} et \ref{eq:rdef}) faisant intervenir le paramètre $\alpha$ qui va devenir clé dans notre étude.
$\alpha$ représente la proportion massique des précurseurs qui sont utilisés pour produire de la machinerie d'expression génique, au détriment de la machinerie métabolique (qui elle est produite à la proportion $1-\alpha$).
Comme on l'a déjà vu avec les lois de croissance établie à l'état stationnaire (voir Chapitre~\ref{chap:introduction}), cette proportion varie avec l'environnement, on s'attend donc à ce que $\alpha$ en fasse autant, augmentant d'autant plus que l'environnement devient riche (Fig.~\ref{fig:scott_rnaprot}).
Cependant, de quelle manière varie-t-il entre deux états de croissance stationnaire, lorsque les composants cellulaires sont en déséquilibre et que la cellule doit adapter sa composition ?

Comme nous l'avons vu dans le Chapitre~\ref{chap:introduction}, les lois de croissances à l'état stationnaire s'expliquent facilement si l'on considère que la cellule maximise son taux de croissance.
En Section~\ref{sec:growthlaws}, nous voyons en effet que pour chaque environnement, un $\alpha$ unique existe qui maximise la croissance de la cellule, mais surtout que ce $\alpha^*_{opt}$ est d'autant plus important que l'environnement est riche (Fig.~\ref{fig:model_analysis}).
De fait, les lois de croissances à l'état stationnaire sont parfaitement prédites par notre modèle simple en considérant uniquement que la cellule maximise son taux de croissance (Fig.~\ref{fig:model_validation}).
Peut-on appliquer le même principe, dans un contexte cette fois dynamique, de façon à prédire comment la cellule distribue ses ressources lors d'une transition de croissance ?

Dans ce contexte, $\alpha$ est désormais une fonction du temps.
Nous cherchons donc quelle fonction du temps maximise un objectif donné.
Ce problème, formulé en Section~\ref{sec:optimalcontrolproblem}, est un problème de contrôle optimal, puisque l'objectif que l'on se fixe est une fonction d'une fonction, plus communément appelée fonctionnelle.
Nous choisissons comme équivalent dynamique de la maximisation du taux de croissance à l'état stationnaire, la maximisation de la production de biomasse sur un intervalle de temps comprenant une changement abrupt de l'environnement (Eq.~\ref{eq:biomassdyn}).
Nous résolvons ce problème en Section~\ref{sec:optimalcontrolsolution} en couplant utilisation du principe du maximum de Pontryagin avec de l'optimisation numérique.
La solution optimale obtenue montre que le contrôle $\alpha$ doit prendre alternativement les valeurs 0 ou 1 jusqu'à ce que le système soit stabilisé sur un nouvel équilibre (Fig.~\ref{fig:optimalcontrol} et Eq.~\ref{eq:optcontrol}).
En d'autres termes, la meilleure production de biomasse s'obtient si, à tout moment, la cellule aiguille 100\% de ses ressources vers un seul secteur, en alternant entre l'un et l'autre jusqu'à ce que le nouvel état stationnaire de croissance soit atteint.
Ce type de solution se rencontre souvent en contrôle optimal, est répond au nom de concept TOR (Tout-Ou-Rien, ou \textit{bang-bang-control} en anglais).
Mais ce contrôle est-il pertinent d'un point de vue biologique ?
La cellule a-t-elle suffisamment d'information à sa disposition pour réaliser une telle transition, laquelle repose sur des changements abrupts de l'expression génique à des moments clés de la transition ?

En Section~\ref{sec:steadystate}, nous utilisons le problème de contrôle optimal et sa solution comme un banc d'essai permettant de comparer entre elle différentes stratégies de régulations (Fig.~\ref{fig:strategies_overview}).
En se servant du modèle pour guider notre intuition, et en nous limitant aux solutions qui respectent les lois de croissance stationnaires, nous montrons que deux stratégies de régulations simples sont possibles : soit l'organisme mesure directement la quantité de nutriments dans l'environnement, soit il mesure la concentration en précurseurs dans la cellule.
De manière intéressante, ces deux stratégies sont optimales à l'état stationnaire, et donc sont exactement équivalentes.
Cependant, dans le contexte dynamique de notre banc d'essai, la stratégie qui consiste à mesurer les précurseurs est bien plus efficace, même si elle reste évidemment inférieure à la solution optimale (Fig.~\ref{fig:comparison_nutrients_precursor}).
Mais il s'agit là de régulations simples qui exploitent l'information ne venant que d'une seule variable.

Finalement, nous montrons en Section~\ref{sec:strategies} qu'une stratégie mesurant à la fois les précurseurs et la concentration en machinerie d'expression génique est tout à fait capable de réaliser un contrôle proche du tout-ou-rien optimal (Eq.~\ref{eq:stratswitch} et Fig.~\ref{fig:onoffresults}).
Mais surtout, en réutilisant un modèle du système ppGpp qui contrôle la synthèse des ribosomes chez la bactérie \textit{Escherichia coli}, nous montrons que ce système répond, au moins structurellement, aux exigences d'une telle stratégie de régulation (Fig.~\ref{fig:ppGppsurface}).
Cela procure un nouveau point de vue sur ce système largement répandu chez de nombreux microorganismes: en intégrant l'information provenant à la fois de la quantité de ribosomes et celle des précurseurs, il permet d'allouer efficacement les ressources de la cellule en environnement variable.
\end{chapter_summary}
\selectlanguage{english}

\begin{center}
\noindent\rule{4cm}{0.1pt}
\end{center}

\section*{Beginning of Chapter \thechapter}

\noindent \textbf{Important note:} Text and figures in this chapter have been published in Plos Computational Biology~\cite{giordano_dynamical_2016} under the terms of the \href{https://creativecommons.org/licenses/by/4.0/}{Creative Commons Attribution License} (CC-BY 4.0).

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=0.7]{./Fig/Fig1}
\caption{
\textbf{Self-replicator model of bacterial growth.}
External substrates $S$ enter the cell and are transformed into precursors $P$ through the action of the metabolic machinery $M$.
The precursors are used by the gene expression machinery $R$ to make the proteins composing both the metabolic machinery (transporters, enzymes, ...) and the gene expression machinery itself (RNA polymerase, ribosomes, ...).
$\alpha$ ($1-\alpha$) is the mass proportion of precursors converted into $R$ ($M$).
Thick arrows denote reactions and thin, dashed arrows denote catalytic activities.
The rate of synthesis of precursors and the rate of synthesis of proteins from precursors are denoted by $V_M$ and $V_R$, respectively.
}
\label{fig:self_replicator}
\end{figure}

@@

\begin{figure}[tb]
\includegraphics[width=\textwidth]{./Fig/Fig2}
\caption{\textbf{Analysis of self-replicator model of bacterial growth.}
\textit{A:}~Phase-plane analysis of the self-replicator model of Eqs~\ref{eq:pdef} and \ref{eq:rdef}.
The nullclines for $p$ and $r$ are shown as solid and dashed curves, respectively.
Parameter values are $e_M = 3.6$~h\textsuperscript{-1}, $k_R = 3.6$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, $\beta = 0.003$~L~g\textsuperscript{-1}, $\alpha = 0.45$.
\textit{B:}~Dependence of the growth rate at steady state $\mu^{*}$ on the resource allocation parameter $\alpha$, for two different environmental conditions (solid line, $e_M = 4.76$~h\textsuperscript{-1}; dashed line, $e_M = 1.57$~h\textsuperscript{-1}, other parameter values are $k_R = 2.23$~h\textsuperscript{-1}, $K_R=1$~g~L\textsuperscript{-1}, and $\beta = 0.003$~L~g\textsuperscript{-1}).
The maximal growth rate is attained for a unique $\alpha$, called $\alpha^*_{opt}$.
}
\label{fig:model_analysis}
\end{figure}

@@

\begin{figure}[p]
\includegraphics[width=\textwidth]{./Fig/Fig3}
\caption{
\textbf{Self-replicator model accounts for bacterial growth laws.}
\textit{A:}~Predicted quasi-linear relation between the maximal growth rate $\mu^{*}_{opt}$ and the corresponding optimal resource allocation $\alpha^*_{opt}$, for different values of $e_M$ (different colors).
The colored dots indicate $\alpha^*_{opt}$ and $\mu^{*}_{opt}$  for $k_R = 2.23$~h\textsuperscript{-1} and different $e_M$, and the dashed black line the relation for all intermediate values of $e_M$.
The dashed colored lines indicate the relation between $\alpha^*_{opt}$ and $\mu^{*}_{opt}$ obtained when, for a given value of $e_M$, the value of $k_R$ is decreased (lower $k_R$ leads to lower $\mu^{*}_{opt}$).
The solid grey curves correspond to $(\mu^*,\alpha)$-profiles like those shown in Fig.~\ref{fig:model_analysis}\textit{B}.
\textit{B:}~Measured relation between the total RNA/protein mass ratio and the growth rate, in different growth media with different doses of a translation inhibitor (data from~\cite{scott_interdependence_2010}).
For each medium, indicated by a color, five different concentrations of inhibitor were used (higher dose leads to lower growth rate).
Growth-medium compositions are given in the original publication and error bars represent standard deviations.
The dashed black and colored lines are the same as in panel \textit{A}, indicating the good quantitative correspondence between model predictions and experimental data for the chosen parameter values, obtained by fitting the model to the data points (see \textit{Methods} for details).
}
\label{fig:model_validation}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig4}
\caption{
\textbf{Optimal control of the self-replicator during a nutrient upshift.}
\textit{A:}~Optimal trajectory in the phase plane for the nondimensionalized model of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, with streamlines. The optimal trajectory is shown as a solid, red curve.
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the switching curve $\varphi(\hat{p})$.
The optimal solution was obtained by numerical optimization using \texttt{bocop}~\cite{bonnans_bocop_2012} (see \textit{Methods} for details), using the parameter values $E_M=1$ and $K=0.003$, and starting from the initial state $(0.024,0.18)$ at $t=0$ (optimal steady state for $E_M=0.2$).
\textit{B:}~Time evolution of the control variable $\alpha_{opt}(\cdot)$ (thick, red line) and the environment $E_M$ (dashed, black line).
}
\label{fig:optimalcontrol}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=\textwidth]{./Fig/Fig5}
\caption{
\textbf{Alternative strategies for controlling the self-replicator of bacterial growth.}
The feedback control strategies, shown in red and superposed on the self-replicator of Fig.~\ref{fig:self_replicator}, exploit information on system variables and the environment to adjust the value of $\alpha$, and thus the relative allocation of resources to the metabolic machinery and gene expression machinery.
}
\label{fig:strategies_overview}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig6}
\caption{
\textbf{Comparison of the performance of the nutrient-only and precursor-only strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the nutrient-only strategy (green curve).
The solid, black curve represents the $\hat{p}$-nullcline.
The dashed, black curve is the $\hat{r}$-nullcline.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with $\alpha = f(E_M)$ as specified by Eq.~\ref{eq:meth_f} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The initial state corresponds to the steady state attained for an environment given by $0.2 \, E_M$.
While converging to the new steady state after the upshift, the precursor concentration makes a large overshoot.
\textit{B:}~As above, but for the precursor-only strategy.
The feedback control strategy is now defined by $\alpha = g(\hat{p})$ as specified by Eq.~\ref{eq:meth_g} in the \textit{Methods} section and plotted in \ref{S1_Fig}.
The solution trajectory (blue curve) exhibits a lower overshoot.
\textit{C:}~Evolution of the control variable $\alpha(\cdot)$ as a function of time, for each of the above two strategies.
Notice that in the nutrient-only strategy $\alpha(\cdot)$ immediately jumps to the optimal value for the post-upshift steady state (green curve), whereas in the precursor-only strategy it depends on the (time-varying) precursor concentration (blue curve).
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$  as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:comparison_nutrients_precursor}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[scale=1]{./Fig/Fig7}
\caption{
\textbf{Comparison of the performance of the precursor-only and the on-off strategies after a nutrient upshift.}
\textit{A:}~Trajectory in the phase plane for the on-off strategy (yellow curve).
The solid, black curve represents the $\hat{p}$-nullcline and the dashed, black curve the function $g$.
The solution is obtained by numerical simulation of the system of Eqs~\ref{eq:pdefnondim}-\ref{eq:rdefnondim}, supplemented with the equation $\alpha = h(\hat{p},\hat{r})$ defined in Eq.~\ref{eq:stratswitch} and plotted in Fig.~\ref{fig:ppGppsurface}\textit{A}.
The initial state corresponds to the optimal steady state attained for an environment given by $0.2 \, E_M$.
\textit{B:}~Trajectory in the phase plane for the precursor-only strategy (same as in Fig.~\ref{fig:comparison_nutrients_precursor}\textit{B}, added for comparison).
\textit{C:}~Evolution of the control variable $\alpha$ for each strategy as a function of time.
Both strategies stabilize the system at the optimal steady state, but only the on-off strategy (yellow curve) exhibits bang-bang behavior.
\textit{D:}~Evolution of the ratio $\texttt{Vol} / \texttt{Vol}_{opt}$ for the on-off and precursor-only strategies as a function of time, where $\texttt{Vol}$ is the volume of the self-replicator and $\texttt{Vol}_{opt}$ the volume of the same replicator following the optimal strategy shown in Fig.~\ref{fig:optimalcontrol}.
The final values of $\texttt{Vol} / \texttt{Vol}_{opt}$ attained by the two strategies are 0.9831 and 0.9413, respectively.
The on-off strategy is thus hardly distinguishable from the optimal control strategy in the plot.
In all of the above simulations, the parameter values $E_M=1$ and $K=0.003$ were used.
}
\label{fig:onoffresults}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[scale=1]{./Fig/Fig8}
\caption{
\textbf{ppGpp regulation implements an on-off control strategy of resource allocation.}
\textit{A:}~Response surface of the on-off control strategy, defined by $\alpha = h(\hat{p},\hat{r})$ in Eq.~\ref{eq:stratswitch}.
\textit{B:}~Response surface of the ppGpp control strategy, as defined by Eq.~\ref{eq:ppGpp_empirical} and the simplified kinetic model defining ppGpp in terms of the total amino acid concentration and the ribosomal protein fraction (\ref{S4_Text}).
The shape of the response surface of the ppGpp control strategy is seen to be in very good agreement with the on-off strategy leading to near-optimal performance of the self-replicator during a nutrient upshift.
}
\label{fig:ppGppsurface}
\end{figure}

@@

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.8\textwidth]{Fig/Fig_sliding}
\caption
{
\textbf{Local stability of the on-off strategy.}
The on-off strategy sets $\alpha$ to a value of 0 (1) when $\hat{r} > g(\hat{p})$ ($\hat{r} < g(\hat{p})$).
The solid, black curve is the $\hat{p}$-nullcline.
The dashed, black curve is the curve $\hat{r}=g(\hat{p})$.
The arrows represent the vector fields for $\alpha=0$ (in blue) and $\alpha=1$ (in red).
The intersection of the $\hat{p}$-nullcline and the curve $\hat{r}=g(\hat{p})$ corresponds to a unique non-trivial stable steady state, which is equal to $(\hat{p}_{opt}^*, \hat{r}_{opt}^*)$ by Eq.~\ref{eq:supp_adim_h}.
}
\label{fig:sliding}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[height=4cm]{Fig/adj2ab} 
\includegraphics[height=4cm]{Fig/adj2} 
\caption{\textbf{Transitions between regions in the phase-plane for the adjoint system.} A switch occurs when a trajectory crosses the $\lambda_p$-axis. Left: abnormal case. An extremal trajectory cannot have more than two switches. Right: normal case. $(\lambda_p^s,0)$ corresponds to the singular arc.  After the first switch, an extremal trajectory cannot have two consecutive switches if it stays in the region $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}<\hat{p}_{opt}^*\}$ or $\{(\hat{p},\hat{r})\in \Omega \mid \hat{p}>\hat{p}_{opt}^*\}$.}
\label{fig-adj}                               
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[width=0.7\textwidth]{./Fig/FigS4-1}
\caption[ppGpp concentration is a function of total ribosome and amino acid concentrations.]
{
{\bf ppGpp concentration is a function of total ribosome and amino acid concentrations.}\newline
We assume the dynamics of ppGpp to be fast on the time-scale of changes in the ribosome and amino acid concentrations.
The concentration of ppGpp can thus be expressed as a function of the latter two variables, using the model of Bosdriesz \textit{et al.}~\cite{bosdriesz_how_2015}.
Parameters are taken from Table~\ref{tab:notations}.
}
\label{fig:ppGpp}
\end{figure}

\begin{table}[p]
\scriptsize
\centering
\begin{tabular}{cccl}
\hline
\hline
\textbf{Symbol} & \textbf{Value} & \textbf{Unit} & \textbf{Description} \\
\hline
\hline
$a_i$ & -- & $\mu$M & Concentration of aa $i$ (not incorporated in protein)\\
